Differential invariant signatures for planar Lie group transformations of images

### 2020-11-16 paper/newpaper.tex
fixes

\begin{document}
\maketitle
\begin{abstract}
\end{abstract}

% REQUIRED
\begin{keywords}

\end{keywords}

@@

\section{Introduction}
%\todo[inline]{
%Richard: 
%Start on invariant tensor theorem
%Generate new pictures, add function details
%SE(2) jobs inline
%Mobius, Conformal -- finish code, tidy writing
%First bash at experimental section

@@

Ideally, an invariant signature would be complete, so that the image is exactly determined up to a transformation, and two objects share a signature if and only if they differ only by a transformation from $\mathcal{G}$. However, this often requires signatures in many dimensions, and potentially also more derivatives of the image. For many applications an incomplete signature is sufficient, and is what we consider here. See, e.g.,~\cite{UsMobius} and references therein for more discussion about the desirable properties of a set of invariants for image and curve analysis.

In this paper we investigate differential invariant signatures for images and parts of images under planar spatial transformations, for both grayscale and colour images. The remainder of this introductory sectin provides a description of the notation that we use in this paper, followed by a summary of relevant work on invariants. We also introduce the way that we will present examples of the signatures for the various groups we consider. Following this, in the next three sections we describe three distinct methods of computing differential invariants: by tensor contraction, using the theory of transvectants, and via the method of moving frames, applying the methods to a variety of the planar Lie groups. We consider the difference that colour images make in Section~\ref{sec:colour}, and then discuss the number of derivatives required and the effect this has on the practicality of the method.

\subsection{Notation}

@@

For the case of two-dimensional greyscale images it is often convenient to work in coordinates. In this case the transformation group action by a coordinate transformation $\varphi \colon \mathbb{R}^2 \to \mathbb{R}^2$ is given by:
\begin{equation}\label{eq:transformation}
  \mathbf{x} \mapsto \bar{\mathbf{x}} = \varphi(\mathbf{x}).
\end{equation}
The transformed image is then defined to be:
\begin{equation}\label{eq:fbarequalsf}
  \bar{f}(\bar{\mathbf{x}}) = (f\circ\varphi^{-1})(\bar{\mathbf{x}}) = f(\mathbf{x}).
\end{equation}
Derivatives of $\bar{f}$ and $f$ can be related by the chain rule, for example the first order derivatives are given by:
\begin{equation}\label{eq:prolongation}
  \begin{aligned}
    f_x &= \bar{f}_{\bar{x}} \bar{x}_x + \bar{f}_{\bar{y}}\bar{y}_x \\
    f_y &= \bar{f}_{\bar{x}} \bar{x}_y + \bar{f}_{\bar{y}}\bar{y}_y \\
  \end{aligned}
\end{equation}
where $\bar{x}_x = \frac{\partial \varphi_1}{\partial x}(x, y)$, et cetera. It is often convenient to work with $\varphi^{-1}$ instead, as in \S\ref{sec:movingframes}.

\subsection{Relevant Literature}
The theory of invariants has a rich mathematical history, see e.g.,~\cite{OlverCIT,Kraft2002} for an overview. Of most relevance to us here 
\cite{Cayley,Littlewood44,Cartan52,Patterson1928,Veblen1872}

@@

---
TEXT FROM OUR MOBIUS PAPER
Calabi et al. [9] propose the use of differential invariant signatures for shape analysis, and further
argue that these should be approximated in a group-invariant way. For example, for the Euclidean
group, the signature is the shape \todo[inline]{something didn't copy paste correctly here} (??; ??s) regarded as a subset of R2. The claimed advantages
of the approach are that the signature determines the shape; that it does not depend on the
choice of initial point on the curve or on parameterization by arclength, and that its G-invariance
makes it robust; and that it is based on a general procedure for arbitrary objects and groups.
See [2, 4, 9, 18, 30, 35, 36] for further developments and applications of the invariant signature.
Although the signature at first sight appears to be complete (e.g., Theorem 5.2 in [9]),
a more detailed treatment (e.g., [17, 28]) highlights the fact that it is not complete on shapes
that contain singular parts { straight and circular segments in the Euclidean case. For these
parts, the signature reduces to a point. Thus, for shapes that have nearly straight or nearly
circular segments, the signature cannot be robust. In addition, while the signature does not
depend on the starting point or the parameterization, it takes values in a very complicated set,
namely, the planar shapes. To compare the invariants of two shapes requires comparing two
shapes. Essentially, the parameterization-dependence has only been deferred to a later stage
of the analysis (unless one is content to compare shapes visually). One approach to this is to
weight the signature, see [32] for more details.
---

@@

JUSTIFY THIS SET OF GROUPS

\todo[inline]{Use T(2) as an example}

\draw[thick,->] (C1) to  (C2);
\draw[thick,->] (C1) to  (C3);
\draw[thick,->] (C1) to  (B1);
\draw[thick,->] (C2) to  (B3);
%\draw[thick,->] (C1) to  (A2);
\draw[thick,->] (C3) to  (A1);
\draw[thick,->] (A1)   to (A2);
\draw[thick,->] (B1) to  (B2);
\draw[thick,->] (B2) to  (B3);
\draw[thick,->] (B2) to  (A2);
\draw[thick,->] (A2) to  (A4);
\draw[thick,->] (B3) to  (A6);
\draw[thick,->] (A4) to  (A6);
\draw[thick,->] (A6) to  (A5);
\end{tikzpicture}
\end{center}
\caption{The lattice of groups considered in this paper. The edges identify subgroups.}
\end{figure}

%Groups we've done:
%translations, E(2), Sim(2), SA(2), A(2), PSL(2,C), PSL(3,R), Diff\_con, Diff\_vol, Diff

We consider as our image, the following function, defined on $[-1, 1]
\times:
[-1, 1]$
\begin{equation}\label{eq:function}
  \begin{split}
    f(x, y) &= 0.5 - 0.2x - 0.3y - 0.05x^2 + 0.03xy + 0.04y^2 \\ 
    &+ 0.03y^3 + 0.001x^3 - 0.0015x^2y + 0.002xy^2 - 0.0005y^3.
  \end{split}
\end{equation}
This function was chosen to have no critical points, and to be
predominantly linear, so that the higher order derivatives were not too
large. This was predominantly so the higher order terms in the signatures
would not overwhelm the smaller terms. The function is depicted as a
greyscale image in Figure~\ref{fig:function}
\begin{figure}
  \centering
  \includegraphics[width=12cm]{Figs/function}
  \caption{The test function used in this document, from
  Eq.~\eqref{eq:function}.}\label{fig:function}
\end{figure}

We consider the signatures of this function under various transformation
groups, as detailed throughout the main body of the paper and illustrate
the signature in three dimensions. A grid is overlaid on the image of the
function itself, as shown in Fig~\ref{fig:function_scanlines}. This grid is
mapped to the signature surface to help understand what the signature looks
like locally.
\begin{figure}
  \centering
  \includegraphics[width=12cm]{Figs/function_scanlines}
  \caption{The test function used in this document, from
  Eq.~\eqref{eq:function} with a grid of lines
overlaid.}\label{fig:function_scanlines}
\end{figure}

\section{Differential invariants through tensor contraction}
While we avoid the use of tensors in most of this study, there is a notable exception to be made for the Euclidean transformation group $E(2)$. In this setting we can restrict ourselves to \emph{Cartesian tensors}, where we need make no distinction between covariance and contravariance as they are equivalent. We therefore use subscript indices exclusively.

@@

While we can construct any number of invariants this way, there is no guarantee of functional independence of the constructed invariants. We will address this issue in Section~\ref{sec:movingframes} when we construct invariants of $E(2)$ and other groups using the method of moving frames.

@@

Another way to compute invariants is to compare the action of group elements on separate copies of the underlying space (here, $\mathbb{R}^2$).

@@

Consider $\mathbf{x}_i \in \mathbb{R}^2, i=1,2$, so that $\mathbf{x}_i$ can
be written in coordinates as $(x_i, y_i)$. Then a pair of such $\mathbf{x}$ lie in $\mathbb{R}^2 \times \mathbb{R}^2$. Now consider applying the same affine transformation to each $\mathbf{x}$ independently: $\mathbf{x}_i \mapsto A \mathbf{x}_i + \mathbf{b}$ where $A \in GL(2,\mathbb{R}$) and $\mathbf{b} \in \mathbb{R}^2$. Then the Omega Process is a second-order differential operator defined by: 
\begin{equation}
\Omega_{ij} = \begin{vmatrix} \frac{\partial}{\partial x_i} &
\frac{\partial}{\partial y_i} \\ \frac{\partial}{\partial x_j} &
\frac{\partial}{\partial y_j} \end{vmatrix} = 
\frac{\partial^2}{\partial x_i
\partial y_j} - \frac{\partial^2}{\partial x_j \partial y_i}.
\end{equation}

Under the above affine transformation, $\Omega_{ij} \mapsto (\det A)^{-1}
\Omega_{ij}$. Note that $(\det A)^{-1}$ has no dependence on the coordinates.
The Omega Process can be applied to pairs of smooth functions as follows,
writing $\frac{\partial f}{\partial x}$ as $f_x$:

@@

Under this formula we end with a factor $(\det A)^{-r}$ in place of $(\det
A)^{-1}$ that we saw previously, when all $n$ copies of $\mathbb{R}^2$ are
transformed by the same affine function $\mathbf{x} \mapsto A\mathbf{x} +
\mathbf{b}$. $r$ is referred to as the weight of the partial
transvectant. Note that the individual pairwise Omega processes commute.

For the current case it is sufficient to consider $n$ copies of the same
function $f$. If we represent each copy (of $\mathbb{R}^2$) as a node on a
undirected graph, and each Omega process $\Omega_{ij}$ as an edge joining
nodes $i$ and $j$, a partial transvectant can be compactly represented in a
graphical form.

For example, consider a weight $2$ transvectant defined as:
\begin{align*}
    \mbox{tr} &\left[(\Omega_{12}\Omega_{12}) f(x_1, y_1)f(x_2, y_2)\right] \\
    &= \mbox{tr}\left[(\partial_{x_1 y_2} - \partial_{x_2y_1})(\partial_{x_1 y_2} -
    \partial_{x_2y_1})f(x_1, y_1)f(x_2, y_2)\right]\\
    &= \mbox{tr}\left[\left(\partial_{x_1 y_2} - \partial_{x_2 y_1}\right)\left(f_x(x_1, y_1)f_y(x_2, y_2) - f_y(x_1, y_1)f_x(x_2, y_2)\right)\right] \\
    &= \mbox{tr}\left[f_{xx}(x_1, y_1)f_{yy}(x_2, y_2) - f_{xy}(x_1, y_1)f_{xy}(x_2, y_2)\right] \\
    &= f_{xx}(x, y) f_{yy}(x, y) - f_{xy}(x, y)^2.
\end{align*}
This is a degree $2$ partial transvectant as it involves derivatives of $f$
up to second-order, and can be represented graphically by the following
diagram:
\begin{center}
\begin{tikzpicture}[baseline=0]
    \node[draw,circle,minimum size=0.2cm] (1) at (-0.5,0) {1};
    \node[draw,circle,minimum size=0.2cm] (2) at (0.5,0) {2};
    \draw[-] (1) to [out=30,in=150] (2);
    \draw[-] (1) to [out=-30,in=-150] (2);
\end{tikzpicture}
\end{center}
where the double edge between nodes $1$ and $2$ corresponds to the two copies of $\Omega_{12}$.
Tables~\ref{tab:transvectants1} and~\ref{tab:transvectants2} shows all possible nonzero partial
transvectants of $A_2$ up to weight $4$ that can be generated in this way.
Note that the weight is the number of edges in the graph.

@@

\todo[inline]{Explain}

@@

Fig.~\ref{fig:SA2} shows the function as a greyscale image, the transformed
function, and the signatures of both the untransformed and transformed
images.
\begin{figure}
\centering
\includegraphics[width=.45\textwidth]{Figs/f_transformed_SA2.png}
\includegraphics[width=.45\textwidth]{Figs/SA2_signature.png}
\caption{\todo[inline]{add extra crap} A sample $SA(2)$ Transform and 3D signature. Mention what's going on with the grid.}
\label{fig:SA2}
\end{figure}

\subsubsection{Moving on to A(2), \ldots}
Using the same function as an example, 
\begin{figure}
\centering
\includegraphics[width=.45\textwidth]{Figs/f_transformed_A2.png}
\includegraphics[width=.45\textwidth]{Figs/A2_signature.png}
\caption{A sample $A(2)$ Transform and 3D signature.}
\label{fig:A2}
\end{figure}

For $A(2)$ the weight of the partial transvectants prevents them from being invariants. They can be normalised by choosing only those where all components have the same weight and projecting them onto a unit hypersphere. Instead, we select ratios of partial transvectants. It is beneficial to choose the numerator and denominator so that they have different critical points, which can be achieved by adding together partial transvectants of the same weight as appropriate.

@@

@@

It's convenient to work in the complex variable $z = x + iy$. Without loss of
generality, we can assume $x = y = 0$. We begin the construction of our
cross section by choosing $\bar{x} = \bar{y} = 0$. Then, $\psi^{-1}$ 
can be represented locally by a Taylor series of the form
\begin{equation*}
    z = \psi^{-1}(\bar{z}) = c_1 \bar{z} + c_2 \bar{z}^2 + c_3\bar{z}^3 + \cdots
\end{equation*}
where $c_k = a_k + ib_k, k = 1, 2, \ldots$. We solve for the parameters
$a_k, b_k$ in stages.  To find $c_1 = a_1 + i b_1$ we need two constraints.
Differentiating $f(x, y) = \bar{f}(\bar{x}, \bar{y})$ and substituting in
our existing cross section constraints $\bar{x} = \bar{y} = 0$ gives
\begin{align*}
    \bar{f}_{\bar{x}} = f_x x_{\bar{x}} + f_y y_{\bar{x}} &= a_1 f_x + b_1
    f_y, \\
    \bar{f}_{\bar{y}} = f_x x_{\bar{y}} + f_y y_{\bar{y}} &= -b_1 f_x + a_1
    f_y. 
\end{align*}
Note that $\bar{x} = \bar{y} = 0$ removes all higher order derivatives from
the cross section equation. Adding two extra constraints to the cross
section, $\bar{f}_{\bar{x}} = 1, \bar{f}_{\bar{y}} = 0$, this system is readily
solved to give 
\begin{equation*}
    \begin{aligned}
        a_1 &= f_x / (f_x^2 + f_y^2), &  b_1 &= f_y / (f_x^2 + f_y^2).
    \end{aligned}
\end{equation*}
We then repeat the process, this time using the second derivative
computations. This time, there are two new parameters, $a_2, b_2$, to solve
for, but there are three derivatives. We use the cross section equations
$\bar{f}_{\bar{x}\bar{x}} = 0, \bar{f}_{\bar{x}\bar{y}} = 0$ to solve for
$a_2$ and $b_2$, the remaining derivative $\bar{f}_{\bar{y}\bar{y}}$ is then the
sole second-order invariant, which we also saw for the M\"obius group: 
\begin{equation*}
    \bar{f}_{\bar{y}\bar{y}} = \frac{f_{xx} + f_{yy}}{f_x^2 + f_y^2}.
\end{equation*}
Continuing in this way, at derivative order $n$ two real group parameters
enter: $a_n$ and $b_n$. There are $n+1$ independent derivatives of $f$ of
order $n$, so there must be $n-1$ new independent invariants at each order.
Hence at third order there are two more invariants, using the constraints
$\bar{f}_{\bar{x}\bar{x}\bar{x}} = 0, \bar{f}_{\bar{x}\bar{x}\bar{y}} = 0$,
the remaining derivatives are invariant:
\begin{align*}
    \bar{f}_{\bar{x}\bar{y}\bar{y}} &= \frac{f_y^3f_{yyy} + f_x^2f_yf_{yyy}
            -2f_y^2f_{yy}^2 - 2f_{xx}f_y^2f_{yy} -
            4f_xf_{xy}f_yf_{yy} - 2f_x^2 f_{xx}f_{yy} + f_{xxy}f_y^3 +
            f_x^2 f_{xxy}f_y - 4 f_x f_{xy} f_{xx} f_y + f_x^3 f_{xxx} -
        2f_x^2 f_{xx}^2 + f_x^3 f_{xyy}}{(f_x^2 + f_y^2)^3} \\
        \bar{f}_{\bar{y}\bar{y}\bar{y}} &= \frac{f_xf_y^2f_{yyy} +
        f_x^3f_{yyy} - 2f_x f_y f_{yy}^2 + 2f_{xy}f_y^2f_{yy} -
    2f_x^2f_{xy}f_{yy} - f_{xxx}f_y^3 - f_{xyy}f_y^3 + f_xf_{xxy}f_y^2 +
2f_{xy}f_{xx}f_y^2 - f_x^2 f_{xxx} f_y + 2f_xf_{xx}^2 f_y - f_x^2 f_{xyy}
f_y + f_x^3 f_{xxy} - 2f_x^2 f_{xy} f_{xx} }{(f_x^2 + f_y^2)^3 }
\end{align*}

@@

\subsection{Volume-Preserving Diffeomorphisms $\mathbf{Diff_{vol}}$}

The subgroup of volume-preserving diffeomorphisms are those diffeomorphisms that preserve the volume form, which in 2D is actually the area form. They can be defined by one function of two variables, the generating function $S$; one possibility is $S(x,y')$ where $(x,y)\mapsto(x',y')$ according to
$$ x'=\frac{\partial S}{\partial y'},\quad y = \frac{\partial S}{\partial x}.$$

Consider expanding $S$ in Taylor series. The linear terms in the map are determined from the quadratic terms in $S$, which contain 3 group parameters, and act on 2 coordinates, $f_x$ and $f_y$.  The quadratic terms in the map are determined from the cubic terms in $S$, which contain 4 group parameters, and act on 3 coordinates, $f_{xx}$, $f_{xy}$, and $f_{yy}$. At order $n$ we have $n+2$ new group parameters acting on $n+1$ coordinates. Thus, barring some exceptional behaviour in the action, we do not expect any invariants at generic points.

This can be confirmed by considering a small image patch with $f(x,y)=x.$ Let $v=\psi_y\partial/\partial_x - \psi_x\partial/\partial_y$ be the generator of an area-preserving map. The image $f$ transforms to $x + \varepsilon \psi_y(x,y)$ which (by choosing $\psi$ appropriately) can be any function close to $x$.

>>>>

\begin{document}
\maketitle
\begin{abstract}
Object recognition, planar transformation, invariants, differential invariant signature, demo of 3 methods with examples for planar Lie groups, discussion of utility
\end{abstract}

% REQUIRED
\begin{keywords}
Object recognition, planar transformation, invariants, differential invariant signature
\end{keywords}

@@

\section{Introduction}
\todo[inline]{
Abstract
Bad set, singular
Literature
Weight

Example intro
T(2)

Sim(2) by contraction
A(2) by transvectant

When introduce each group, mention relevance to images?

Friendly intro to moving frame
Finish colour

Discussion
}

%\todo[inline]{
%Richard: 
%Start on invariant tensor theorem
%Generate new pictures, add function details
%SE(2) jobs inline
%Mobius, Conformal -- finish code, tidy writing
%First bash at experimental section

@@

Ideally, an invariant signature would be complete, so that the image is exactly determined by the signature up to a transformation, and two objects share a signature if and only if they differ only by a transformation from $\mathcal{G}$. However, this often requires signatures in many dimensions, and potentially also more derivatives of the image. For many applications an incomplete signature is sufficient, and is what we consider here. See, e.g.,~\cite{UsMobius} and references therein for more discussion about the desirable properties of a set of invariants for image and curve analysis.

In this paper we investigate differential invariant signatures for images and parts of images under planar spatial transformations, primarily Lie groups, for both greyscale and colour images. The set of groups that we consider are shown as a lattice in Figure~\ref{fig:lattice}, where the size of the groups decrease as down the page, and the arrows denote the subgroup relationship. There is an important relationship between a group and its subgroups for invariants, which is that invariants of a group are automatically invariants of its subgroups. This means that differentiating between them requires care.

The remainder of this introductory section provides a description of the notation that we use in this paper, followed by a summary of relevant work on invariants. We also introduce the way that we will present examples of the signatures for the various groups we consider. Following this, in the next three sections we describe three distinct methods of computing differential invariants: by tensor contraction, using the theory of transvectants, and via the method of moving frames, applying the methods to a variety of the planar Lie groups. We consider the difference that colour images make in Section~\ref{sec:colour}, and then discuss the number of derivatives required and the effect this has on the practicality of the method.

\draw[thick,->] (C1) to  (C2);
\draw[thick,->] (C1) to  (C3);
\draw[thick,->] (C1) to  (B1);
\draw[thick,->] (C2) to  (B3);
%\draw[thick,->] (C1) to  (A2);
\draw[thick,->] (C3) to  (A1);
\draw[thick,->] (A1)   to (A2);
\draw[thick,->] (B1) to  (B2);
\draw[thick,->] (B2) to  (B3);
\draw[thick,->] (B2) to  (A2);
\draw[thick,->] (A2) to  (A4);
\draw[thick,->] (B3) to  (A6);
\draw[thick,->] (A4) to  (A6);
\draw[thick,->] (A6) to  (A5);
\end{tikzpicture}
\end{center}
\caption{The lattice of groups considered in this paper. The edges identify subgroups.}
\label{fig:lattice}
\end{figure}

\subsection{Concepts and Notation}

@@

For the case of two-dimensional greyscale images it is often convenient to work in coordinates. In this case the transformation group action by a coordinate transformation $\varphi \colon \mathbb{R}^2 \to \mathbb{R}^2$ is given by:
\begin{equation}\label{eq:transformation}
  \mathbf{x} \mapsto \bar{\mathbf{x}} = \varphi(\mathbf{x}).
\end{equation}
The transformed image is then defined to be:
\begin{equation}\label{eq:fbarequalsf}
  \bar{f}(\bar{\mathbf{x}}) = (f\circ\varphi^{-1})(\bar{\mathbf{x}}) = f(\mathbf{x}).
\end{equation}
Derivatives of $\bar{f}$ and $f$ can be related by the chain rule, for example the first order derivatives are given by:
\begin{equation}\label{eq:prolongation}
  \begin{aligned}
    f_x &= \bar{f}_{\bar{x}} \bar{x}_x + \bar{f}_{\bar{y}}\bar{y}_x \\
    f_y &= \bar{f}_{\bar{x}} \bar{x}_y + \bar{f}_{\bar{y}}\bar{y}_y \\
  \end{aligned}
\end{equation}
where $\bar{x}_x = \frac{\partial \varphi_1}{\partial x}(x, y)$, et cetera. %It is often convenient to work with $\varphi^{-1}$ instead, as in \S\ref{sec:movingframes}.

\subsection{Relevant Literature}

The theory of invariants has a rich mathematical history, see e.g.,~\cite{OlverCIT,Kraft2002} for an overview. Of most relevance to us here 
\cite{Cayley,Littlewood44,Cartan52,Patterson1928,Veblen1872}

@@

@@

%Groups we've done:
%translations, E(2), Sim(2), SA(2), A(2), PSL(2,C), PSL(3,R), Diff\_con, Diff\_vol, Diff
\subsection{Example signatures}

In this paper our aim is develop the mathematical underpinnings of constructing differential invariant signatures for images under the action of planar Lie groups. However, as we develop the signature for each group, we will provide an example of the type of surface that the signature defines. Note that these will generally be three-dimensional representation of the planar image.

The image we use is defined on $[-1,1] \times [-1,1]$ to be:

\begin{equation}\label{eq:function}
**write me**
\end{equation}

The function is depicted as a greyscale image in Figure~\ref{fig:function}. On the left the image itself is shown, while on the right a version with a regular grid superimposed is given. This is useful for visualising both the transformation and how the image is mapped into the three-dimensional signature surface.

\begin{figure}
  \centering
  \includegraphics[width=.45\textwidth]{Figs/function}
  \includegraphics[width=.45\textwidth]{Figs/function_scanlines}
  \caption{The test function used in this document, from
  Eq.~\eqref{eq:function}.}\label{fig:function}
\end{figure}

\subsection{Getting Started: Translations}

The most basic transformation group of a image is the translation group $T(2): \phi(\mathbf{x}) = \mathbf{x} + \mathbf{t}$.

We could explicitly write the calculation in Eq.~\ref{eq:prolongation} for this case, but instead provide an intuitive explanation. Consider a point in the original image, and the corresponding point in the transformed one. Clearly, the intensity value of the points match, and so do their first spatial derivatives. These three quantities can be used as coordinates in a new space, where the two image points will correspond. Computing these coordinates for all image points will define a surface in the new three-dimensional space, and the two images will produce the same surface, which we call the signature.

Figure~\ref{fig:T2} shows (a) the original image, (b) the translated image by $x,y$, (c) the signature of one of them, (d) the superimposed pair of signatures. Now explain the missing bits.

\section{Differential invariants through tensor contraction}

While we avoid the use of tensors in most of this study, there is a notable exception to be made for the Euclidean transformation group $E(2)$. In this setting we can restrict ourselves to \emph{Cartesian tensors}, where we need make no distinction between covariance and contravariance as they are equivalent. We therefore use subscript indices exclusively.

@@

While we can construct any number of invariants this way, there is no guarantee of functional independence between them. We will address this issue in Section~\ref{sec:movingframes} when we construct invariants of $E(2)$ and other groups using the method of moving frames.

@@

Another way to compute invariants is to compare the simultaneous action of a group element on separate copies of the underlying space (here, $\mathbb{R}^2$). By identifying these spaces with each other it is possible to construct a function known as a transvectant that is invariant up to a scale factor (a \emph{relative invariant}) to that group action.

@@

Consider $\mathbf{x}_i \in \mathbb{R}^2, i=1,2$, so that $\mathbf{x}_i$ can
be written in coordinates as $(x_i, y_i)$; a pair of such $\mathbf{x}$ lie in $\mathbb{R}^2 \times \mathbb{R}^2$. Now consider applying the same affine transformation to each $\mathbf{x}$ independently: $\mathbf{x}_i \mapsto A \mathbf{x}_i + \mathbf{b}$ where $A \in GL(2,\mathbb{R}$) and $\mathbf{b} \in \mathbb{R}^2$. Then the Omega Process is a second-order differential operator defined by: 
\begin{equation}
\Omega_{ij} = \begin{vmatrix} \frac{\partial}{\partial x_i} &
\frac{\partial}{\partial y_i} \\ \frac{\partial}{\partial x_j} &
\frac{\partial}{\partial y_j} \end{vmatrix} = 
\frac{\partial^2}{\partial x_i
\partial y_j} - \frac{\partial^2}{\partial x_j \partial y_i}.
\end{equation}

Under the above affine transformation, $\Omega_{ij} \mapsto (\det A)^{-1}
\Omega_{ij}$, and is thus a relative invariant. Note that $(\det A)^{-1}$ has no dependence on the coordinates.
The Omega Process can be applied to products of pairs of smooth functions as follows,
writing $\frac{\partial f}{\partial x}$ as $f_x$:

@@

Under this formula we end with a factor $(\det A)^{-r}$ in place of $(\det
A)^{-1}$ that we saw previously, when all $n$ copies of $\mathbb{R}^2$ are
transformed by the same affine function $\mathbf{x} \mapsto A\mathbf{x} +
\mathbf{b}$. $r$ is referred to as the weight of the partial
transvectant. Note that the individual pairwise Omega processes commute.
The concept of a relative invariant will be important in what follows.

For the current case it is sufficient to consider $n$ copies of the same
function $f$. If we represent each copy of $\mathbb{R}^2$ as a node on a
undirected graph, and each Omega process $\Omega_{ij}$ as an edge joining
nodes $i$ and $j$, a partial transvectant can be compactly represented in a
graphical form.

For example, consider a weight $2$ transvectant defined as:
\begin{align*}
    \mbox{tr} &\left[(\Omega_{12}\Omega_{12}) f(x_1, y_1)f(x_2, y_2)\right] \\
    &= \mbox{tr}\left[(\partial_{x_1 y_2} - \partial_{x_2y_1})(\partial_{x_1 y_2} -
    \partial_{x_2y_1})f(x_1, y_1)f(x_2, y_2)\right]\\
    &= \mbox{tr}\left[\left(\partial_{x_1 y_2} - \partial_{x_2 y_1}\right)\left(f_x(x_1, y_1)f_y(x_2, y_2) - f_y(x_1, y_1)f_x(x_2, y_2)\right)\right] \\
    &= \mbox{tr} 2 \left[f_{xx}(x_1, y_1)f_{yy}(x_2, y_2) - f_{xy}(x_1, y_1)f_{xy}(x_2, y_2)\right] \\
    &= 2 f_{xx}(x, y) f_{yy}(x, y) - 2 f_{xy}(x, y)^2.
\end{align*}
This is a degree $2$ partial transvectant as it involves derivatives of $f$
up to second-order, and can be represented graphically by the following
diagram:
\begin{center}
\begin{tikzpicture}[baseline=0]
    \node[draw,circle,minimum size=0.2cm] (1) at (-0.5,0) {1};
    \node[draw,circle,minimum size=0.2cm] (2) at (0.5,0) {2};
    \draw[-] (1) to [out=30,in=150] (2);
    \draw[-] (1) to [out=-30,in=-150] (2);
\end{tikzpicture}
\end{center}
where the double edge between nodes $1$ and $2$ corresponds to the two copies of $\Omega_{12}$ (the names are purely for convenience).

Tables~\ref{tab:transvectants1} and~\ref{tab:transvectants2} shows all possible nonzero partial
transvectants of $A_2$ up to weight $4$ that can be generated in this way.
Note that the weight is the number of edges in the graph.

@@

@@

Fig.~\ref{fig:SA2} shows the function as a greyscale image, the transformed
function, and the signatures of both the untransformed and transformed
images.
\begin{figure}
\centering
\includegraphics[width=.45\textwidth]{Figs/f_transformed_SA2.png}
\includegraphics[width=.45\textwidth]{Figs/SA2_signature.png}
\caption{A sample $SA(2)$ transformation and 3D signature.}
\label{fig:SA2}
\end{figure}

\subsubsection{Moving on to A(2), \ldots}
Using the same function as an example, 
\begin{figure}
\centering
\includegraphics[width=.45\textwidth]{Figs/f_transformed_A2.png}
\includegraphics[width=.45\textwidth]{Figs/A2_signature.png}
\caption{A sample $A(2)$ transformation and 3D signature.}
\label{fig:A2}
\end{figure}

For $A(2)$ the weight of the partial transvectants prevents them from being invariants. There are two options to create a signature. One is to select ratios of partial transvectants. It is then beneficial to choose the numerator and denominator so that they have different critical points, which can be achieved by adding together partial transvectants of the same weight as appropriate. Instead, we normalise the partial transvectants by choosing only those where all components have the same weight and projecting them onto the unit hypersphere. By multiplying these by the function (which takes values between 0 and 1) we create a signature enclosed by the unit hypersphere. An example is shown in Figure~\ref{fig:A2}.

@@

** Given the importance of the conformal camera model ...

@@

It is convenient to work in the complex variable $z = x + iy$. Without loss of
generality, we can assume $x = y = 0$. We begin the construction of our
cross section by choosing $\bar{x} = \bar{y} = 0$. Then, $\psi^{-1}$ 
can be represented locally by a Taylor series of the form:
\begin{equation*}
    z = \psi^{-1}(\bar{z}) = c_1 \bar{z} + c_2 \bar{z}^2 + c_3\bar{z}^3 + \cdots,
\end{equation*}
where $c_k = a_k + ib_k, k = 1, 2, \ldots$. We solve for the parameters
$a_k, b_k$ in stages.  To find $c_1 = a_1 + i b_1$ we need two constraints.
Differentiating $f(x, y) = \bar{f}(\bar{x}, \bar{y})$ and substituting in
our existing cross section constraints $\bar{x} = \bar{y} = 0$ gives:
\begin{align*}
    \bar{f}_{\bar{x}} = f_x x_{\bar{x}} + f_y y_{\bar{x}} &= a_1 f_x + b_1
    f_y, \\
    \bar{f}_{\bar{y}} = f_x x_{\bar{y}} + f_y y_{\bar{y}} &= -b_1 f_x + a_1
    f_y. 
\end{align*}
Note that $\bar{x} = \bar{y} = 0$ removes all higher order derivatives from
the cross section equation. Adding two extra constraints to the cross
section, $\bar{f}_{\bar{x}} = 1, \bar{f}_{\bar{y}} = 0$, this system is readily
solved to give :
\begin{equation*}
    \begin{aligned}
        a_1 &= f_x / (f_x^2 + f_y^2), &  b_1 &= f_y / (f_x^2 + f_y^2).
    \end{aligned}
\end{equation*}
We then repeat the process, this time using the second derivative
computations. This time, there are two new parameters, $a_2, b_2$, to solve
for, but there are three derivatives. We use the cross section equations
$\bar{f}_{\bar{x}\bar{x}} = 0, \bar{f}_{\bar{x}\bar{y}} = 0$ to solve for
$a_2$ and $b_2$, the remaining derivative $\bar{f}_{\bar{y}\bar{y}}$ is then the
sole second-order invariant, which we also saw for the M\"obius group: 
\begin{equation*}
    \bar{f}_{\bar{y}\bar{y}} = \frac{f_{xx} + f_{yy}}{f_x^2 + f_y^2}.
\end{equation*}
Continuing in this way, at derivative order $n$ two real group parameters
enter: $a_n$ and $b_n$. There are $n+1$ independent derivatives of $f$ of
order $n$, so there must be $n-1$ new independent invariants at each order.
Hence at third order there are two more invariants, using the constraints
$\bar{f}_{\bar{x}\bar{x}\bar{x}} = 0, \bar{f}_{\bar{x}\bar{x}\bar{y}} = 0$,
the remaining derivatives are invariant:
\begin{align*}
    \bar{f}_{\bar{x}\bar{y}\bar{y}} &= \frac{f_y^3f_{yyy} + f_x^2f_yf_{yyy}
            -2f_y^2f_{yy}^2 - 2f_{xx}f_y^2f_{yy} -
            4f_xf_{xy}f_yf_{yy} - 2f_x^2 f_{xx}f_{yy} + f_{xxy}f_y^3 +
            f_x^2 f_{xxy}f_y - 4 f_x f_{xy} f_{xx} f_y + f_x^3 f_{xxx} -
        2f_x^2 f_{xx}^2 + f_x^3 f_{xyy}}{(f_x^2 + f_y^2)^3} \\
        \bar{f}_{\bar{y}\bar{y}\bar{y}} &= \frac{f_xf_y^2f_{yyy} +
        f_x^3f_{yyy} - 2f_x f_y f_{yy}^2 + 2f_{xy}f_y^2f_{yy} -
    2f_x^2f_{xy}f_{yy} - f_{xxx}f_y^3 - f_{xyy}f_y^3 + f_xf_{xxy}f_y^2 +
2f_{xy}f_{xx}f_y^2 - f_x^2 f_{xxx} f_y + 2f_xf_{xx}^2 f_y - f_x^2 f_{xyy}
f_y + f_x^3 f_{xxy} - 2f_x^2 f_{xy} f_{xx} }{(f_x^2 + f_y^2)^3 }
\end{align*}

@@
